An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-19 paper.tex
Added comments on Sec. 3.

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
These neighborhoods should be non-overlapping and fill parameter
space.  To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  In fact, the proposed
jumps are drawn from a piecewise-constant (i.e.\ constant on each
neighborhood) interpolation of the PDF.  There are various techniques
that could be used to construct the set of neighborhoods associated
with each point.

@@

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively:
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to Step 1 with each of the sub-boxes, storing the
  resulting trees as sub-trees of the current box.
\end{enumerate}
The key algorithmic step in the production of a kD-tree is finding the
median point along a given dimension in order to divide the points in
half in Step 2.  For $n$ points, this can be accomplished in
$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
points in total, there are $\order{\log N}$ levels in the tree; at
each level, $\order{N}$ points must be processed once in the
median-finding algorithm.  Tree construction thus costs $\order{N \log
  N}$ in time, and the tree consumes $\order{N \log N}$ space.  Box
boundaries for a kD-tree constructed around a point set that is
normally distributed around the origin in two dimensions are shown in
Figure \ref{fig:kD-tree}.

@@

In order to use the kD-tree interpolation in an MCMC, we must be able
to quickly find the neighborhood associated with a given point.  This
can be accomplished in $\order{\log N}$ time and constant space with
the following algorithm.  Given the point, $\vtheta_i$---which could
either be a point in the chain of MCMC samples if we are trying to
propose a jump, or a previously proposed jump if we are trying to
compute the jump probability distribution (see
Eq.~\ref{eq:p-accept})---and the tree, $T$:
\begin{enumerate}
\item If $T$ contains exactly one sample point, then its box is the
  associated neighborhood.  Otherwise:
\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
  contained in the ``left'' sub-tree, then return to Step 1,
  considering this sub-tree; otherwise return to Step 1, considering
  the ``right'' sub-tree.
\end{enumerate}

>>>>

%\ilya{I think this section could just be phrased in terms of interpolation rather than jump proposal distributions; i.e., "the problem of finding the local density" instead of "the problem of drawing a proposed jump", etc.}

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
These neighborhoods should be non-overlapping and fill parameter
space.  \ilya{I don't think this is generic enough: for example, Littenberg and Cornish also constructed an interpolation, but one in which multiple points could be in the same "neighborhood", while other neighborhoods could be empty. I.e., the algorithm below is one way of solving this problem, but not an alternative statement of it.}
To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF. \ilya{This last statement is presumably only true if the size of the neighborhood associated with each point is roughly the inverse of the local point density.  I think it's worthwhile to point that out that your kD trees allow you to just that in a computationally efficient manner.}  In fact, the proposed
jumps are drawn from a piecewise-constant (i.e.\ constant on each
neighborhood) interpolation of the PDF.  There are various techniques
that could be used to construct the set of neighborhoods associated
with each point.

@@

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively:
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to Step 1 with each of the sub-boxes, storing the
  resulting trees as sub-trees of the current box.
\end{enumerate}
The key algorithmic step in the production of a kD-tree is finding the
median point along a given dimension in order to divide the points in
half in Step 2.  For $n$ points, this can be accomplished in
$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
points in total, there are $\order{\log N}$ levels in the tree; at
each level, $\order{N}$ points must be processed once in the
median-finding algorithm.  Tree construction thus costs $\order{N \log
  N}$ in time, and the tree consumes $\order{N \log N}$ space.  As an illustration, box
boundaries for a kD-tree constructed around a point set that is
normally distributed around the origin in two dimensions are shown in
Figure \ref{fig:kD-tree}.

@@

In order to use the kD-tree interpolation in an MCMC, we must be able
to quickly find the neighborhood associated with a given point.  This
can be accomplished in $\order{\log N}$ time and constant space with
the following algorithm.  Given the point, $\vtheta_i$ --- which could
either be a point in the chain of MCMC samples if we are trying to
propose a jump, or a previously proposed jump if we are trying to
compute the jump probability distribution (see
Eq.~\ref{eq:p-accept}) --- and the tree, $T$:
\begin{enumerate}
\item If $T$ contains exactly one sample point, then its box is the
  associated neighborhood.  Otherwise:
\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
  contained in the ``left'' sub-tree, then return to Step 1,
  considering this sub-tree; otherwise return to Step 1, considering
  the ``right'' sub-tree.
\end{enumerate}
